---
title: "Bermuda Atlantic Time Series-Ocean acidification"
author: "Leira J. Centeno"
date: '2022-04-25'
output: 
  html_document:
   code_folding: "show" 
  
---
**Introduction**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


#Load library

library(tidyverse)
library(seacarb)
library(performance)

#import BATS Bottle data from website

bats_bottle <- read_table2("http://batsftp.bios.edu/BATS/bottle/bats_bottle.txt", 
                           col_names = FALSE, skip = 60)
#View(bats_bottle)

#assign column column names to BATS data from website
colnames(bats_bottle)=
  colnames(read_csv("http://batsftp.bios.edu/BATS/bottle/bats_bottle.txt", skip = 59))
#View(bats_bottle)

#Varible names and units form BATS data
#yyyymmdd = Year Month Day   
#decy   = Decimal Year     
#time   = Time (hhmm)      
#latN   = Latitude (Deg N) 
#lonW   = Longitude (Deg W)
#Depth  = Depth (m)                  
#Temp   = Temperature ITS-90 (C)    
#CTD_S  = CTD Salinity (PSS-78)      
#Sal1   = Salinity-1 (PSS-78)        
#Sig-th = Sigma-Theta (kg/m^3)       
#O2(1)  = Oxygen-1 (umol/kg)          
#OxFixT = Oxygen Fix Temp (C)        
#Anom1  = Oxy Anomaly-1 (umol/kg)    
#CO2    = dissolved inorganic carbon (umol/kg)              
#Alk    = Alkalinity (uequiv)        
#NO3    = Nitrate+Nitrite-1 (umol/kg)
#NO2    = Nitrite-1 (umol/kg)        
#PO4    = Phosphate-1 (umol/kg)      
#Si     = Silicate-1 (umol/kg)       
#POC    = POC (ug/kg)                
#PON    = PON (ug/kg)                
#TOC    = TOC (umol/kg)                
#TN     = TN (umol/kg)  NOTE: Prior to BATS 121, DON is reported instead of TON
#Bact   = Bacteria enumeration (cells*10^8/kg)
#POP    = POP (umol/kg)
#TDP    = Total dissolved Phosphorus (nmol/kg)
#SRP    = Low-level phosphorus (nmol/kg)
#BSi    = Particulate biogenic silica (umol/kg)
#LSi    = Particulate lithogenic silica  (umol/kg)
#Pro    = Prochlorococcus (cells/ml)
#Syn    = Synechococcus (cells/ml)
#Piceu  = Picoeukaryotes (cells/ml)
#Naneu  = Nanoeukaryotes (cells/ml)   

?carb
#carb(flag, var1, var2, S=35, T=25, Patm=1, P=0, Pt=0, Sit=0,
#     k1k2="x", kf="x", ks="d", pHscale="T", b="u74", gas="potential", 
#     warn="y", eos="eos80", long=1.e20, lat=1.e20)
#

#we're missing pressure but we can calculate this from depth using TEOS-10 toolbox
bats_bottle= 
  bats_bottle %>% 
  mutate(sea_pressure=gsw_p_from_z(Depth*-1,latN))

#now we have all the variables we need to calculate carbonate chemistry!
#but we must be VERY Careful about our units and matching the correct variables

bats_co2=
bats_bottle %>% 
rename(DIC=CO2) %>% 
  filter(Depth<10) %>% 
  filter(Alk!=-999,DIC!=-999,Sal1!=-999,Temp!=-999,sea_pressure!=-999,
         PO41!=-999,Sal1!=-999,lonW!=-99,latN!=-999) %>%
mutate(carb(flag=15, var1=Alk*10^-6, var2=DIC*10^-6, S=Sal1, T=Temp, Patm=1, P=sea_pressure/10, Pt=PO41*10^-6, Sit=Si1*10^-6,
   k1k2="l", kf="x", ks="d", pHscale="T", b="u74", gas="potential", 
    warn="y", eos="teos10", long=lonW, lat=latN))

#view(bats_co2)
```
## 1) Is surface ocean pCO2 increasing?
```{r,warning = FALSE, message = FALSE}
bats_co2 %>% 
  ggplot()+
  geom_point(mapping = aes(x=decy, y=pCO2))+
  geom_smooth(mapping = aes(x=decy, y=pCO2), method="lm")+
  ylab(" pCO2") +
  xlab("Years")+
  coord_cartesian(xlim = c(1991,2020))+
 annotate("text",x=2000, y= 445,label="R^2=0.17", size=6, color="red")+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor= element_blank(),
        panel.background = element_blank(), #removes annoying gray area
        axis.line= element_line(colour = "black"), #adds lines for the axes
        text = element_text(size = 20))
```

### How to cuantify if its increasing?  
```{r}
pco2_model= with(bats_co2, lm(pCO2~I(decy-2000)))
```

```{r}
                                    data <- anova(pco2_model)
                                    knitr::kable(data,
                                               caption = "Anova for pCO2")
```
```{r}
#DUDAAAAA
summary(pco2_model)
check_model(pco2_model)
```

p-value=2.2e-16 there's significance difference. 

## 2) Is surface ocean seawater pH decreasing? 
```{r, warning = FALSE, message = FALSE}
  #PLOT
bats_co2 %>% 
  ggplot()+
  geom_point(mapping = aes(x=decy, y=pH))+
  geom_smooth(mapping = aes(x=decy, y=pH), method="lm")+
    ylab(" pH") +
    xlab("Years")+
  coord_cartesian(xlim = c(1991,2020))+
    theme(panel.grid.major = element_blank(), 
          panel.grid.minor= element_blank(),
          panel.background = element_blank(), #removes annoying gray area
          axis.line= element_line(colour = "black"), #adds lines for the axes
          text = element_text(size = 20))

```

### How to cuantify if its increasing? 
```{r}
#How to cuantify if its increasing? statistics 
pH_model= with(bats_co2, lm(pH~I(decy-2000)))
```

```{r}
 data <- anova(pH_model)
                          knitr::kable(data,
                                  caption = "Anova for pH")
```

```{r}
summary(pH_model)
check_model(pH_model)
```

p-value = 1.461e-15 there's significance difference.

## 3a) Is surface ocean omega decreasing?
```{r,warning = FALSE, message = FALSE}
#ARAGONITE
bats_co2 %>% 
  ggplot()+
  geom_point(mapping = aes(x=decy, y=OmegaAragonite))+
  geom_smooth(mapping = aes(x=decy, y=OmegaAragonite), method="lm") +
  ylab("Omega Aragonite") +
  xlab("Years")+
  coord_cartesian(xlim = c(1991,2020))+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor= element_blank(),
        panel.background = element_blank(), #removes annoying gray area
        axis.line= element_line(colour = "black"), #adds lines for the axes
        text = element_text(size = 20))
```
### How to cuantify if its increasing? 
```{r}
OmeAragonite_model= with(bats_co2, lm(OmegaAragonite~decy))
```

```{r}
 data <- anova(OmeAragonite_model)
                          knitr::kable(data,
                                  caption = "Anova for Omega Aragonite")
```

```{r}
summary(OmeAragonite_model)
check_model(OmeAragonite_model)
```
p-value= 1.353e-08

## 3b) Is surface ocean omega decreasing?
```{r, warning = FALSE, message = FALSE}
#Calcite
bats_co2 %>% 
  ggplot()+
  geom_point(mapping = aes(x=decy, y=OmegaCalcite))+
  geom_smooth(mapping = aes(x=decy, y=OmegaCalcite, method="lm")) +
  ylab(" Omega Calcite") +
  xlab("Years")+
  coord_cartesian(xlim = c(1991,2020))+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor= element_blank(),
        panel.background = element_blank(), #removes annoying gray area
        axis.line= element_line(colour = "black"), #adds lines for the axes
        text = element_text(size = 20))
```

### How to cuantify if its increasing? 
```{r}
OmeCalcite_model= with(bats_co2, lm(OmegaCalcite~decy))
```

```{r}
 data <- anova(OmeCalcite_model)
                          knitr::kable(data,
                                  caption = "Anova for Omega Calcite")
```

```{r}
summary(OmeCalcite_model)
check_model(OmeCalcite_model)
```

p-value= 9.619e-12